Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-12-29 demystify.tex
updates

% \chapter{Fellowship Applications}\label{sec:fellowship}
% \chapterinfo{How to apply for fellowships such as NSF GRFP, DOD DSEG to fund your PhD.}

@@

% \paragraph{Evaluating the application's research potential, \emph{not} the project potential} This is very different than a normal proposal like an NSF one that asks for funding for a specific project. In such case the project is evaluated.  However, in the GRFP, the research potential of the applicant is evaluated, not the project.  This means that you should focus on your research experience, your potential to do research, and how you can contribute to the field, etc.  You should also focus on your broader impacts, e.g.,  how you can help improve STEM education through outreach and mentoring, how you can contribute to diversity and inclusion in STEM fields, how you can help society, etc.

>>>>

\chapter{Fellowship Applications}\label{sec:fellowship}
\chapterinfo{How to apply for fellowships such as NSF GRFP, DOD DSEG to fund your PhD.}

@@

\section{How Applications are Evaluated}

\begin{itemize}
\item Each applicant is assigned to ~3 reviewers who have expertise in the applicant's research topics (so only your field but also your specific area).  This means reviewers can be quite knowledgeable in your field and your work, your LoR writers and their work, and thus can evaluate your application quite well.

\item While we review applicants from various levels (e.g., undergrad, 1st year, 2nd year in PhD program), we evaluate and compare applicants \emph{within} their cohort level.  This means that we do not compare an undergrad, who is still applying to grad school, to a 2nd year PhD student who already has an adviser and published multiple top papers. This ensures that each applicant is evaluated fairly based on their level.

\section{Intellectual Merit and Broader Impacts}
Your will be evaluated based on two criteria: Intellectual Merit (IM) and Broader Impacts (BI).

\paragraph{Intellectual merit} is based on your research potential, e.g., your research experience, your potential to do research, and how you can contribute to the field.

Note that unlike a normal project or research proposal, in which the evaluation is based on the project, in the GRFP, the evaluation is \emph{not} about the research plan. We are repeatedly reminded to not focus too much on the research plan as many applicants are still applying for PhD programs and might not even get into the field or work with a prof. that fit their research plan.  Instead, the evaluation is based on your \emph{potential} of being a good researcher using various criteria, e.g., your research experience, what others think of you (LoRs), research plan, etc. This also means that even if you have very little research experience, you can still get the fellowship if you can convince the reviewers that you have the potential to be a good researcher.  Vice versa, if you have a lot of research experience and in very top places but cannot convince the reviewers that you have the potential to be a good researcher, you will not get the fellowship.

\paragraph{Broader impacts} based on how you and your research can benefit society.  Many students just mention how their research is cross-discipline and thus has broader impacts (e.g., improve software quality and thus help society).  This is too common and not enough.  You also want to talk about how you, as a grad student with your research, can help society, e.g., through outreach, mentoring, diversity, etc. As mentioned below, you will want to be specific and concrete, e.g., ``I will work with highschool students and underrepresented students'' is not as convincing as ``I have had experience in mentoring highschool students X and will continue to do so through Y,Z ...''.

\section{Common Pitfalls}

\paragraph{Too technical} Remember that you're writing to a general audience in CS, not experts in your field.  While NSF tries to match reviewers with your field, CS fields are simply too broad and you might get a reviewer in the field but not in your specific area. 
So avoid technical jargon and explain things in simple terms. If the reviewer does not understand your proposal, they will likely give you a low score.

You should ask your professor or someone who's familiar with writing NSF proposals to review your reseach plan.  They can tell you if it is too technical and help you simplify it.

\paragraph{Preliminary Work} Your research plan should have some preliminary work to convince reviewers that you have thought about the problem and have some initial promising results.  It does not have to a lot, e.g., you don't need published results or an implementation, a small experiment done by hand on some small examples would suffice.  Moreover, it should be concrete and convincing, e.g., \emph{``When being applied to the program in Fig 1. of the paper published PLDI'19 that approach X failed, my idea worked and was able to ...''}.  This will set you apart from others who just have a plan but no results.

% Prelim work does not, and in fact should not, be extensive. You do not need a paper or a complete implementation.  You don't need to show that your idea is better than others or even that it will work.  You just need to show that you have tried it and it has some promise.  This can be as simple as a small experiment done by hand on some small examples,  e.g., if you come up with a new abstraction in PL for proving correctness, then you can just manually apply it to a few small programs, preferrably ones that existing works have trouble with.  This is enough to show that you have thought about the problem and have some initial results.

% To make prelim concrete (and thus more convincing), you might want to use specific names and numbers, e.g.,  \emph{``I tried my prototype on the MNIST dataset, achieving an accuracy of 90\%''}, or \emph{``I used Lean to formalize and prove a specific case of the Fermat's Last Theorem''}.  In addition, negative results are results and can be very convincing.  So don't be afraid to say \emph{``I tried my idea on X but it failed because of Y, and I plan to address this by Z''}.

\paragraph{Research} is the emphasis.  Many made a mistake of focusing on teaching and mentoring and not on research. This is especially true for students who have a lot of teaching and mentoring experience, e.g., from a teaching-focused university or a teaching assistant.  While teaching and mentoring are important, the GRFP is a \emph{research} fellowship which aims to train researchers. So you will need to demonstrate your passion for research and your potential to be a good researcher.
